---
title: "Machine learning project"
author: "Rosa"
date: "6/4/2021"
output: html_document
---

# **Background**

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

# **Data**

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

What you should submit
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

Peer Review Portion
Your submission for the Peer Review portion should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).

Course Project Prediction Quiz Portion
Apply your machine learning algorithm to the 20 test cases available in the test data above and submit your predictions in appropriate format to the Course Project Prediction Quiz for automated grading.

# **Analysis**

## Loading data

First of all we start reading and exploring the data

```{r}
library(data.table)
library(dplyr)
library(caret)
#read and explore data
data <- fread("projecte/pml-training.csv")
str(data)
```

## **Cross validation**

We create the train data and the test data so we can do a cross validation of our model
```{r}
set.seed(4657)
all_data_train <- createDataPartition(y=data$classe,p=.70,list=F)
data_training <- data[all_data_train,]
data_testing <- data[-all_data_train,]
```

## **Data cleaning**
When exploring the data, we realized there are few variables that will not be used to create our model, since it would not make sense, these variables are: identifier, user_name, time_stamp. Moreover we have detected many variables with a lot of NA values, these ones will alse be removed so do not interfere in the model.

```{r}
#removing variables cannot use for training: identifier, user_name, time_stamp

data_training <- data_training[,-c(1:7)]

#removing variables with all cases NAs

data_training <- data_training %>% 
  select_if(~!all(is.na(.)))

#select variables with high (over 95%) missing data --> exclude them from the analysis
data_training[data_training==""] <- NA
NArate <- apply(data_training, 2, function(x) sum(is.na(x)))/nrow(data_training)
NArate <- data.frame(as.list(NArate))
NArate_names <- colnames(NArate %>% select_if(~any(. <= 0.95)))
data_training <- data_training %>% select(NArate_names)
```

## **PCA**

We have many variables available and we try to reduce them using a PCA. We have seen than keeping 25 components we are able to explain 95% of the variance.

```{r}
pca_preproc <- preProcess(data_training[,1:52],method="pca",thresh=0.8)
pca_preproc <- preProcess(data_training[,1:52],method="pca",thresh=0.95)

pca_preproc <- preProcess(data_training[,1:52],method="pca",pcaComp = 25)
data_training_pc <- predict(pca_preproc, data_training[,1:52])
```

## **Random forest**
Finally, we use the components resulting from the PCA as variables for the final model. 
```{r}
library(randomForest)
data_training$classe <- factor(data_training$classe)
mod_fit_rf <- randomForest(data_training$classe ~ .,data=data_training_pc, do.trace=F)
print(mod_fit_rf) # view results 
importance(mod_fit_rf)
```

## **Check with test data set**
```{r}
data_testing <-  data_testing[,-c(1:7)]
data_testing[data_testing==""] <- NA
NArate <- apply(data_testing, 2, function(x) sum(is.na(x)))/nrow(data_testing)
NArate <- data.frame(as.list(NArate))
NArate_names <- colnames(NArate %>% select_if(~any(. <= 0.95)))
data_testing <- data_testing %>% select(NArate_names)
data_testing$classe <-  factor(data_testing$classe)
data_testing_pc <- predict(pca_preproc,data_testing[,1:52])
confusionMatrix(data_testing$classe,predict(mod_fit_rf,data_testing_pc))
```

## **Predict classes of 20 cases**
```{r}
final_pred <-fread("projecte/pml-testing.csv")
final_pred <- final_pred[,-(1:7)]
final_pred[final_pred==""] <- NA
NArate <- apply(final_pred, 2, function(x) sum(is.na(x)))/nrow(final_pred)
NArate <- data.frame(as.list(NArate))
NArate_names <- colnames(NArate %>% select_if(~any(. <= 0.95)))
final_pred <- final_pred %>% select(NArate_names)
final_pred_PC <- predict(pca_preproc,final_pred[,1:52])
final_pred$classe <- predict(mod_fit_rf,final_pred_PC)
```

## **Summary**
For this analysis, 19622 observations from weight lifting exercise were used to analyze and predict correct body movement from others during the exercise. Data was seperate in two differents sets, one for building and training the model (70% of the obbservations) and the rest of observations were used for testing the model (cross-validation). The model used was a random forest, previously using a PCA to summarise the data.  The model statistics showed the overall accuracy of 97% for the testing set. 

